Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2011-08-25 rjournal/AE.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@62 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@64 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Having obtained the test statistic, the $p$-value must then be calculated. 
For larger sample sizes, the null distribution
of the test statistic can be approximated
with the null distribution from the classical Kolmogorov-Smirnov test. 
When an exact $p$-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used in
for one-sided tests.  For two-sided tests, the methods presented in
\citet{gleser85} lead to exact two-sided $p$-values.  This requires the
calculation of rectangular probabilities for uniform order statistics
as discussed by \citet{nieder81}.
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in the papers of
Conover and Gleser.

@@

We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most $p$-values, while
this upper bound is used for smaller $p$-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).  Figure \ref{cvmissues} shows this bound with the blue dashed
line; values in red exceeding the bound are a result of the numerical
instability.  Although it would be preferable to determine the use of
the bound based on values of the test statistic rather than the $p$-value,
the range of ``extreme'' values of the test statistic varies with the
hypothesized distribution.

@@

>>>>

Having obtained the test statistic, the $p$-value must then be calculated. 
When an exact $p$-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used in
for one-sided tests.  For two-sided tests, the methods presented in
\citet{gleser85} lead to exact two-sided $p$-values.  This requires the
calculation of rectangular probabilities for uniform order statistics
as discussed by \citet{nieder81}.
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in the papers of
Conover and Gleser.

For larger sample sizes, the classical Kolmogorov-Smirnov test is used
and is known to produce conservative p-values for discrete distributions;
the revised \code{ks.test()} supports estimation
of p-values via simulation if desired.

@@

We resolve this problem by using a combination of
two simple conservative approximations to
avoid the numerical instability. First, consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \label{ineq1} \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\label{ineq2}
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom.  Second,
consider the Markov inequality: 
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{E}
  \left[\exp\left(t\sum_{i=1}^{p}\lambda_iZ_i^2\right)\right]\exp(-tx) \\
&= \frac{\exp(-tx)}{\sqrt{\prod_{i=1}^{p}(1-2t\lambda_i)}}
\label{ineq3}
\end{align}
where the bound can be
minimized over $t \in (0,1/2\lambda_{max})$.
The upper bounds for the $p$-value given by (\ref{ineq2}) and (\ref{ineq3})
are both calculated and the smaller is used in cases where the numerical
instability of (\ref{distQ}) is a concern.

The original formulation, numerical integration of (\ref{distQ}),
 is preferable for most $p$-values, while
the upper bound described above
is used for smaller $p$-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation).  Figure \ref{cvmissues} shows the bounds with the blue and green
dashed lines; values in red exceeding the bounds are a result of the numerical
instability.  Although it would be preferable to determine the use of
the bound based on values of the test statistic rather than the $p$-value,
the range of ``extreme'' values of the test statistic varies with the
hypothesized distribution.

@@

NOTE: Add an example or two with simulated p-values.  Remember to add
stats:: everywhere appropriate for clarity, and ks.test::, etc...
